---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r}
setwd("C:/RWD")
unzip(zipfile="repdata-data-activity.zip")
activity <- read.csv("activity.csv")

```

## What is mean total number of steps taken per day?

```{r}
total.steps <- tapply(activity$steps, activity$date, FUN=sum, na.rm=TRUE)
hist(total.steps, main = paste("Histogram of Total Number of Steps Taken per Day"), xlab="Sum of Steps")
print(paste0("Mean of the total number of steps taken per day is : ",mean(total.steps, na.rm=TRUE)))
print(paste0("Median  of the total number of steps taken per day is : ",median(total.steps, na.rm=TRUE)))
```

## What is the average daily activity pattern?

```{r}
library(ggplot2)
avgStepsInt <- tapply(activity$steps,activity$interval,mean, na.rm=TRUE)
plot(names(avgStepsInt), avgStepsInt, type="l", main = "Time Series Plot", xlab="5-minute Intervals", ylab="Avg Steps")
print(paste0("Interval ", as.numeric(names(which.max(avgStepsInt))) , " contains the maximum number of steps."))
```

## Imputing missing values

```{r}
print(paste0("There are ", sum(is.na(activity)) , " missing values in the activity dataset."))
```

 To fill all of the missing values in the dataset they are going to be replaced with the mean value for that 5-minute interval.

Create a copy of the dataset
```{r}
activityNoNA <- activity
```

There are 288 5-minute periods per day and 61 days. We already have the mean per period in **avgStepsInt** for the 288 period. So lets now we will multiply that by 61 so we can compare to **activityNoNA** and replace.

```{r}
allMeanSteps <- rep(avgStepsInt, 61)
```

The steps before replacement

```{r}
rawSteps <- activityNoNA$steps
```

Find any values that are NA in the raw steps data

```{r}
stepsNA <- is.na(rawSteps)
```

Now replace these values with their corresponding mean

```{r}
rawSteps[stepsNA] <- allMeanSteps[stepsNA]
```

Now replace in our new data set

```{r}
activityNoNA$steps <- rawSteps
```

Get the sums of the new data that we will use in the histogram

```{r}
newTotalSteps <- tapply(activityNoNA$steps,activityNoNA$date,sum, na.rm=TRUE)
```

Okay now we get the new Histogram and the mean/medians

```{r}
hist(newTotalSteps, main = paste("Histogram of Total Number of Steps Taken per Day"), xlab="Sum of Steps")
```

now we calculate the new mean/medians

```{r}
meanSteps <- mean(newTotalSteps)
medianSteps <- median(newTotalSteps)
print(paste0("Mean total number of steps taken per day is: ", meanSteps))
print(paste0("Median total number of steps taken per day is: ", medianSteps))
```

The mean and median are higher and equal. The histogram is higher as a result of showing a summed figure for each interval and we replaced NA's with values. 

## Are there differences in activity patterns between weekdays and weekends?

I have commented the code below. Basically we just loop through the dataset and see if the interval was on a weekend or weekday

```{r}
#Create a vector to compare to for the weekend days
weekend <- c("Sunday","Saturday")
#this we use to store where the interval was during a weekday or weekend
dayWeek <- c()
#Now we check which and store
for (i in 1:length(activityNoNA$steps))
    {
    if (any(weekdays(as.Date(activityNoNA[i,2])) == weekend)) 
        {
            #it was a weekend
            dayWeek[i] <- "weekend"  }
            #a weekday
        else { dayWeek[i] <- "weekday"
         }
    }
## Bind to our dataset
    activityNoNA <- cbind(activityNoNA,dayWeek)
```

Add the library lattice and reshape

```{r}
library(lattice)
library(reshape)
```

Now we group by interval and dayweek and get the mean for steps. Then melt to make it easier to plot then plot.

```{r}
data <- tapply(activityNoNA$steps,list(activityNoNA$interval,activityNoNA$dayWeek),mean)
data <- melt(data)
xyplot(value ~ X1 | X2 ,data=data,panel=panel.lines,layout=c(1,2),xlab="TimeScale",ylab="Steps")
```